---
title: "Biome classification workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Biome classification workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(biomes)
```

# Goal

This vignette shows how to classify species occurrence data into
biome schemes shipped with the `biomes` package. The workflow has
three building blocks:

- `biomes_get()` — load the 31 biome raster layers.
- `biomes_info()` / `biomes_information` — inspect biome metadata.
- `biomes_classify()` — assign each occurrence to a biome.
- `biomes_biome_tab()` — tabulate occurrences per biome.

---

# 1. Load biome layers and metadata

```{r}
layers <- biomes_get()
layers
```

`layers` is a `SpatRaster` with 31 layers. Each layer corresponds to one
row of the `biomes_information` data frame:

```{r}
data(biomes_information)
biomes_information[1, c("publication", "name_of_classification",
                        "layer_in_raster_stack")]
```

For an interactive summary of a specific scheme:

```{r, eval=FALSE}
biomes_info(1)
```

---

# 2. Classify occurrence records

`biomes_classify()` accepts a `data.frame` (with longitude/latitude
columns), an `sf` object, or a `terra::SpatVector`. By default it uses
all 31 layers.

```{r}
data(biomes_example)

# Classify against a single layer, return biome names
class_1 <- biomes_classify(
  x     = biomes_example,
  biome = layers[[1]],
  value = "name"
)
head(class_1)
```

Use `value = "ID"` for the raw raster value, or `value = "both"` to keep
both side by side. With multiple layers, one `_value` / `_name` column
is produced per layer:

```{r}
class_multi <- biomes_classify(
  x     = biomes_example,
  biome = layers[[c(1, 5, 25)]],
  value = "both"
)
head(class_multi)
```

For raster values that have no entry in the legend (azonal classes
encoded with high values such as 90+), the name column falls back to
`"azonal (raster value: X)"` rather than `NA`.

---

# 3. Tabulate occurrences per biome

`biomes_biome_tab()` summarizes the number of **occurrence records**
(one row of the input data = one occurrence) per biome and layer:

```{r}
class_multi |>
  biomes_biome_tab()
```

For unique-species counts per biome, deduplicate by species before
tabulating (see the *Species per biome* article).

---

# 4. Custom coordinate columns

If your data uses different longitude/latitude column names, pass them
via `lon` and `lat`:

```{r}
biomes_classify(
  x     = biomes_example,
  biome = layers[[1]],
  lon   = "decimalLongitude",
  lat   = "decimalLatitude",
  value = "name"
) |>
  head(3)
```

---

# Summary

The typical workflow is:

1. Load layers with `biomes_get()` and pick the scheme(s) of interest.
2. Inspect them with `biomes_info()` or `biomes_information`.
3. Classify your occurrences with `biomes_classify()`.
4. Summarize with `biomes_biome_tab()` and / or join the classification
   back to your data for further analysis.
