---
title: "Biome classification workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Biome classification workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = TRUE,
  warning = TRUE
)
library(biomes)
```

# Goal

This vignette shows how to classify species occurrence data into biome schemes.  
Two common use cases:

- You only know a taxon name.  
  You want to download occurrences from GBIF and classify them.

- You already have your own cleaned dataset (longitude, latitude, taxon column).  
  You want to classify it directly.

## Key functions

- `biomes_GBIF()`  
  Download and clean GBIF occurrences for a given taxon.

- `biomes_get()`  
  Load the packaged biome raster stack and legend into your session.

- `biomes_compare()`  
  High-level helper. Runs biome assignment, summaries, and plotting for one or more biome layers and writes the result objects into your environment.

- `biomes_count()` / `biomes_plot()`  
  Lower-level building blocks. `biomes_compare()` calls these per layer.


All examples below are shown with `eval = FALSE` so they do not run during CRAN checks.

---

# 1. Download GBIF data for a taxon

`biomes_GBIF()` is a convenience wrapper around rgbif. By default it uses `occ_search()` so you do not need GBIF credentials. It filters for records with coordinates and (optionally) cleans obvious coordinate issues.

```{r eval=FALSE}
occ_raw <- biomes_GBIF(
  taxon = "Solemyida",
  use_download = FALSE,     # default pathway (occ_search)
  filter_clean = TRUE,      # basic CoordinateCleaner filters
  filter_sea = FALSE,       # optionally remove marine-only points
  year_min = 2000,
  year_max = 2024,
  limit = 5000
)

dim(occ_raw)
head(occ_raw[, c("scientificName",
                 "decimalLongitude",
                 "decimalLatitude")])
```

Minimum required columns for downstream steps:
- decimalLongitude
- decimalLatitude
- a taxon column such as scientificName or family

These columns will later be mapped to biome classes.

---

# 2. Load biome data

`biomes_get()` loads two core objects from the package into your environment:

- biome_raster: A SpatRaster containing all biome layers.
- biome_legend: A table describing each layer (layer index, source, biome class labels, etc.).

It can also (optionally) open static pre-rendered PNG maps of specific layers.

```{r eval=FALSE}
# Load biome_raster and biome_legend into the global environment
biomes_get()

# Optionally: show quick-look maps for specific layers (if shipped PNGs exist)
biomes_get(view_layer = c(1, 5))
```

After running `biomes_get()`, you should see biome_raster and biome_legend in your workspace.
You can inspect which biome layers exist like this:

```{r eval=FALSE}
terra::nlyr(biome_raster)
head(biome_legend)
```

---

# 3. Compare biome schemes

`biomes_compare()` is the high-level entry point for typical users. It will:

- Harmonize your occurrence data (or download from GBIF if you only gave taxon).
- Decide which biome layer(s) to use.
- Run `biomes_count()` and `biomes_plot()` for each layer.
- Store the results in the target environment.

## Case A: Only a taxon name

You do not have a dataset yet. You just want to say "classify this taxon".

```{r eval=FALSE}
biomes_compare(
  taxon = "Solemyida",
  layer = NULL,            # let the function decide which biome layer(s)
  biomes_suggest = TRUE,   # call biomes_suggest() to pick a 'best' layer
  env = .GlobalEnv         # write output objects into the global environment
)
```

What this does internally:
- Calls `biomes_GBIF("Solemyida")`.
- Cleans and standardizes column names. It creates a column `species` because downstream code expects a column with that name.
- Picks biome layers. If `layer = NULL` and `biomes_suggest = TRUE`, it will call `biomes_suggest()` and use its chosen layer.
- For each chosen layer it runs:
    - `biomes_count()` to assign each point to a biome class and summarize per-biome presence.
    - `biomes_plot()` to generate a map of occurrences per cell and barplots of total occurrences and unique taxa per layer.

After it finishes, the following objects appear in `.GlobalEnv`:

- `biomes_compare_input`: The occurrence data actually used in the analysis. Includes `species`.
- `biomes_compare_layers`: Integer vector of biome layer indices that were evaluated.
- `biomes_compare_summary`: One row per biome layer. Contains total number of mapped occurrences, number of unique taxa, and source info for that biome scheme.
- `biomes_count_<L>`: Output of `biomes_count()` for layer L.
- `biomes_plot_<L>`: Output of `biomes_plot()` for layer L.

You can inspect them:

```{r eval=FALSE}
biomes_compare_layers
biomes_compare_summary

names(biomes_count_1)
biomes_count_1$counts_per_biome

names(biomes_plot_1)
biomes_plot_1$barplots$stats_df
biomes_plot_1$barplots$p_occ
biomes_plot_1$barplots$p_sp
```

---

## Case B: You already have your own dataset

Assume you have a cleaned data frame `df` with these columns:

- decimalLongitude
- decimalLatitude
- scientificName (or family, genus, etc.)

Example dummy data:

```{r eval=FALSE}
df <- data.frame(
  decimalLongitude = c(10.1, 10.2, 10.3, -60.5, -60.6),
  decimalLatitude  = c(50.0, 50.1, 50.2,  -3.2,  -3.3),
  scientificName   = c("Taxon_A", "Taxon_A", "Taxon_B", "Taxon_C", "Taxon_C")
)

df
```

Now run `biomes_compare()` directly on your data. No GBIF call will be made.

```{r eval=FALSE}
biomes_compare(
  x = df,
  lon = "decimalLongitude",
  lat = "decimalLatitude",
  taxon_col = "scientificName",
  layer = c(1, 5, 10),     # test multiple biome schemes
  biomes_suggest = FALSE,  # skip automatic suggestion
  env = .GlobalEnv
)
```

Again you will get:

- biomes_compare_input
- biomes_compare_layers
- biomes_compare_summary

plus one `biomes_count_<layer>` and one `biomes_plot_<layer>` object per tested layer.

---

# 4. Understanding `biomes_count()` output

Each `biomes_count_<L>` object (the result from `biomes_count()`) contains:

- `counts_per_biome`: One row per biome class in that layer. Includes:
    - number of unique taxa considered "present" in that biome,
    - total number of occurrence records in that biome.

- `summary_species_biome`: A presence matrix (taxon × biome).
    - Value 1 means that taxon is considered present in that biome.
    - Presence is controlled by two arguments in `biomes_count()`:
        - `presence_min_n`: minimum number of records in a biome.
        - `presence_min_prop`: optional minimum fraction of that taxon's records that must fall in that biome.

Default is `presence_min_n = 1`, meaning a taxon is counted as present in a biome if it has at least one record there.

You can call `biomes_count()` manually and adjust these thresholds:

```{r eval=FALSE}
res_strict <- biomes_count(
  x = biomes_compare_input,
  species = "species",          # column created by biomes_compare()
  lon = "decimalLongitude",
  lat = "decimalLatitude",
  layer = 5,
  presence_min_n = 5,           # require at least 5 records
  presence_min_prop = 0.1,      # AND at least 10% of that taxon's records
  summary = TRUE,
  return_format = "augment"
)

res_strict$counts_per_biome
res_strict$summary_species_biome
```

You can also ask `biomes_count()` to write CSVs by giving the `path` argument:

```{r eval=FALSE}
res_export <- biomes_count(
  x = biomes_compare_input,
  species = "species",
  lon = "decimalLongitude",
  lat = "decimalLatitude",
  layer = 5,
  path = "results/layer5_export",   # will create several CSV files
  summary = TRUE
)
```

This writes:
- layer5_export_biom.csv
- layer5_export_summary_species_biome.csv
- layer5_export_counts_per_biome.csv

---

# 5. Understanding `biomes_plot()` output

Each `biomes_plot_<L>` object (result from `biomes_plot()`) contains two kinds of products:

- Per-layer occurrence maps  
  Raster cells are colored by how many occurrences fall inside.
  Biome boundaries are overlaid.

- Summary barplots  
  - total number of occurrence records per layer
  - total number of unique taxa per layer

You can access the barplot data and ggplot objects:

```{r eval=FALSE}
# Barplot data frame
biomes_plot_5$barplots$stats_df

# Barplot of total occurrences per layer
biomes_plot_5$barplots$p_occ

# You can save any ggplot:
ggplot2::ggsave(
  filename = "total_occurrences_per_layer.png",
  plot = biomes_plot_5$barplots$p_occ,
  width = 6,
  height = 4,
  dpi = 300
)

# Map for layer 5
biomes_plot_5$map_results[["5"]]$plot
```

---

# 6. Choosing a biome scheme

Different biome layers represent different biome concepts (for example remotely sensed land cover vs. potential natural vegetation vs. climate-driven biome definitions).

`biomes_compare_summary` gives you, per layer:
- total number of occurrences mapped
- number of unique taxa detected
- source label from the legend

This helps you justify and report which biome scheme you will use:

```{r eval=FALSE}
biomes_compare_summary
```

If you ran `biomes_compare(..., biomes_suggest = TRUE)`, then `biomes_compare_layers` typically has just one entry. That entry is the layer suggested by `biomes_suggest()`, which tries to match:
- spatial coverage of your points
- reasonable number of biome classes in your region
- the conceptual basis (for example climate vs. land cover)

---

# Summary

This workflow:
- Obtain or provide occurrence data (`biomes_GBIF()` or your own data frame).
- Load biome layers and metadata (`biomes_get()`).
- Compare biome schemes (`biomes_compare()`).
- Inspect and export summaries (`biomes_count()`, `biomes_plot()`).

The result is a reproducible link from occurrences to biome classes, with both tables and publication-ready figures.
