---
title: "compare_biome_definitions"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(biomes)
```


## Compare bime definitions for specific datasets

You may want to compare the differences of two biome definitions for a given 
occurrence dataset. For two definitions a levelplot based on the number of occurrences
can show the number of overlapp between different biomes between both deifnitions.

```{r}
library(ggplot)
library(biomes)
library(viridis)

data(biomes_example)
layers <- biomes_get()

class <- biomes_classify(x = biomes_example,
                         biome = layers[[c(1,25)]])

tab <- table(t[, names(t)[2]],
             t[, names(t)[3]]) %>% 
  data.frame()


ggplot()+
  geom_raster(data = tab,
              aes(x = Var1,
                  y = Var2,
                  fill = log(Freq)))+
  scale_fill_viridis()+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, 
                                   vjust = 0.5, 
                                   hjust = 1))
```
For comparison of multiple biome definitions a multivariant analyses, for instance a PCoA can 
help to identify clusters of similar biomes accross definitions.

```{r}
library(vegan)
library(tidyverse)

# occurrence to biome classificaiton
class <- biomes_classify(x = biomes_example,
                         biome = layers[[c(1,12, 25)]])

# get number of occurrences per species per biome
test <- biomes_example %>%
  bind_cols(class)%>%
  select(species, contains("Biome_Inventory")) %>%
  pivot_longer(cols = contains("Biome_Inventory"),
               names_to = "layer",
               values_to = "value") %>%
  group_by(layer, value, species) %>%
  count() %>%
  pivot_wider(id_cols = c(value,layer),
              names_from = species,
              values_from = n)

# preapre data for ordination
dat <- test[,-c(1:2)]
dat[is.na(dat)] <- 0

# ordination
ord <- metaMDS(dat, trace = FALSE)
plo <- data.frame(cbind(ord$points,
                        test[,c(1,2)]))

# visualize first 2 axes
ggplot()+
  geom_text(data = plo,
             aes(x = MDS1,
                 y = MDS2,
                 label = value,
                 color = layer))+
  theme_bw()

```
