---
title: "Compare biome definitions"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

```{r setup}
library(biomes)
library(ggplot2)
library(viridis)
library(dplyr)
library(tidyr)
```


## Compare biome definitions for a specific dataset

You may want to compare two biome definitions for a given occurrence
dataset. For two definitions a tile plot based on the number of
occurrences shows how biome classes overlap between the two schemes.

```{r}
data(biomes_example)
layers <- biomes_get()

class <- biomes_classify(
  x     = biomes_example,
  biome = layers[[c(1, 25)]]
)

# class has one *_name column per layer
name_cols <- grep("_name$", names(class), value = TRUE)
tab <- as.data.frame(
  table(class[[name_cols[1]]], class[[name_cols[2]]])
)
names(tab) <- c("biome_layer_1", "biome_layer_2", "Freq")

ggplot(tab) +
  geom_raster(aes(x    = biome_layer_1,
                  y    = biome_layer_2,
                  fill = log1p(Freq))) +
  scale_fill_viridis() +
  labs(x = name_cols[1], y = name_cols[2]) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5,
                                   hjust = 1))
```

For comparing multiple biome definitions a multivariate ordination
(e.g. NMDS) can help identify clusters of similar biomes across
definitions.

```{r, eval = requireNamespace("vegan", quietly = TRUE)}
library(vegan)

class <- biomes_classify(
  x     = biomes_example,
  biome = layers[[c(1, 12, 25)]]
)

# number of occurrences per species per biome, wide format
test <- biomes_example |>
  bind_cols(class) |>
  select(species, ends_with("_name")) |>
  pivot_longer(cols      = ends_with("_name"),
               names_to  = "layer",
               values_to = "biome") |>
  group_by(layer, biome, species) |>
  count() |>
  pivot_wider(id_cols     = c(biome, layer),
              names_from  = species,
              values_from = n)

# prepare data for ordination
dat <- test[, -c(1:2)]
dat[is.na(dat)] <- 0

# ordination
ord <- metaMDS(dat, trace = FALSE)
plo <- data.frame(cbind(ord$points, test[, c(1, 2)]))

ggplot(plo, aes(x = MDS1, y = MDS2, label = biome, color = layer)) +
  geom_text() +
  theme_bw()
```
